---
title:  "NC 2026 Primary:<br>Unaffiliated Voter Ballot Choice by County"
subtitle: "Which party ballot did unaffiliated voters choose?"
date: today
format:
  html:
    theme: flatly
    toc: true
    embed-resources: true
    fig-align: center
    include-in-header:
      - text: |
          <style>
          .quarto-title-block .quarto-title {
            text-align: center;
          }
          .quarto-title-block .quarto-title .title,
          .quarto-title-block .quarto-title .subtitle,
          .quarto-title-block .quarto-title .quarto-categories {
            text-align: center;
          }
          #title-block-header.quarto-title-block {
            text-align: center;
          }
          </style>
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(sf)
library(tigris)
library(scales)
library(stringr)

options(tigris_use_cache = TRUE)
```

## Data

```{r load-data}
# Adjust the object name or file path to match your setup
# absentee <- read_tsv("absentee_20260303.csv")
absentee <- read.csv("~/Downloads/absentee_20260303 2.csv")
```

```{r wrangle}
una_ballots <- absentee %>%
  filter(
    voter_party_code == "UNA",
    ballot_rtn_status == "ACCEPTED",
    ballot_request_party %in% c("DEM", "REP")
  ) %>%
  count(county_desc, ballot_request_party, name = "n") %>%
  pivot_wider(
    names_from  = ballot_request_party,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(
    total   = DEM + REP,
    pct_dem = DEM / total * 100,
    pct_rep = REP / total * 100,
    margin  = pct_dem - pct_rep,
    county_upper = toupper(trimws(county_desc))
  )

total_una   <- sum(una_ballots$total)
total_dem   <- sum(una_ballots$DEM)
total_rep   <- sum(una_ballots$REP)
pct_dem_all <- total_dem / total_una * 100
pct_rep_all <- total_rep / total_una * 100
```

The number of unaffiliated voters in North Carolina has been steadily growing, raising the question of how likely these voters are to disrupt voting patterns and electoral outcomes in the state. Because North Carolina has semi-closed primaries, unaffiliated voters can choose which primary they want to vote in by requesting either the Democratic or Republican ballot. While some voters may make strategic choices — for example, an unaffiliated voter who leans Democratic picking the Republican ballot to support the candidate least likely to win the general election — this number is likely small. Examining ballot choice in primaries is therefore a useful starting point for understanding the inclinations of unaffiliated voters. Here, I provide a first cut by looking at whether ballot choice is related to the partisan leaning of NC counties.

In the March 3, 2026 primary, **`r comma(total_una)`** unaffiliated voters cast either a Democratic or Republican ballot. Statewide, **`r sprintf("%.1f%%", pct_dem_all)`** chose the Democratic ballot and **`r sprintf("%.1f%%", pct_rep_all)`** chose the Republican ballot.

```{r shapefile}
nc_counties <- counties(state = "NC", cb = TRUE, year = 2022) %>%
  mutate(county_upper = toupper(NAME))

nc_map <- nc_counties %>%
  left_join(una_ballots, by = "county_upper")
```

```{r theme}
theme_nc_map <- function() {
  theme_void(base_family = "Helvetica") +
    theme(
      plot.background   = element_rect(fill = "white", color = NA),
      panel.background  = element_rect(fill = "white", color = NA),
      plot.title        = element_text(color = "#1a1a1a", size = 16, face = "bold",
                                       hjust = 0.5, margin = margin(t = 10, b = 4)),
      plot.subtitle     = element_text(color = "#555555", size = 10, hjust = 0.5,
                                       margin = margin(b = 12)),
      plot.caption      = element_text(color = "#888888", size = 7, hjust = 0.5,
                                       margin = margin(t = 10)),
      legend.position   = "bottom",
      legend.title      = element_text(color = "#333333", size = 9),
      legend.text       = element_text(color = "#555555", size = 8),
      legend.key.width  = unit(2, "cm"),
      legend.key.height = unit(0.3, "cm"),
      plot.margin       = margin(10, 15, 10, 15)
    )
}
```

## Percent Choosing the Democratic Ballot

```{r map-dem}
#| fig-width: 10
#| fig-height: 6

ggplot(nc_map) +
  geom_sf(aes(fill = pct_dem), color = "white", linewidth = 0.25) +
  scale_fill_gradient(
    low      = "#deebf7",
    high     = "#084594",
    na.value = "#e8e8e8",
    name     = "% Chose DEM Ballot",
    labels   = function(x) paste0(round(x), "%"),
    guide    = guide_colorbar(title.position = "top", title.hjust = 0.5)
  ) +
  labs(
    title   = "Unaffiliated Voters Choosing the Democratic Ballot",
    subtitle = "NC 2026 Primary Election — by County",
    caption = "Source: NC State Board of Elections absentee data (March 3, 2026 primary)"
  ) +
  theme_nc_map()
```

## Percent Choosing the Republican Ballot

```{r map-rep}
#| fig-width: 10
#| fig-height: 6

ggplot(nc_map) +
  geom_sf(aes(fill = pct_rep), color = "white", linewidth = 0.25) +
  scale_fill_gradient(
    low      = "#fee0d2",
    high     = "#a50f15",
    na.value = "#e8e8e8",
    name     = "% Chose REP Ballot",
    labels   = function(x) paste0(round(x), "%"),
    guide    = guide_colorbar(title.position = "top", title.hjust = 0.5)
  ) +
  labs(
    title    = "Unaffiliated Voters Choosing the Republican Ballot",
    subtitle = "NC 2026 Primary Election — by County",
    caption  = "Source: NC State Board of Elections absentee data (March 3, 2026 primary)"
  ) +
  theme_nc_map()
```

## Democratic vs Republican Margin

```{r map-margin}
#| fig-width: 10
#| fig-height: 6

max_abs_margin <- max(abs(nc_map$margin), na.rm = TRUE)
margin_limit   <- ceiling(max_abs_margin / 10) * 10

ggplot(nc_map) +
  geom_sf(aes(fill = margin), color = "white", linewidth = 0.25) +
  scale_fill_gradient2(
    low      = "#b2182b",
    mid      = "#f7f7f7",
    high     = "#2166ac",
    midpoint = 0,
    limits   = c(-margin_limit, margin_limit),
    na.value = "#e8e8e8",
    name     = "DEM ← Margin → REP",
    labels   = function(x) {
      ifelse(x > 0, paste0("D+", abs(round(x))),
             ifelse(x < 0, paste0("R+", abs(round(x))), "Even"))
    },
    guide = guide_colorbar(title.position = "top", title.hjust = 0.5, reverse = TRUE)
  ) +
  labs(
    title    = "Which Party Won Unaffiliated Voters?",
    subtitle = "NC 2026 Primary — DEM vs REP ballot margin among UNA voters, by County",
    caption  = "Source: NC State Board of Elections absentee data (March 3, 2026 primary)"
  ) +
  theme_nc_map()
```


::: {style="text-align: center;"}
## Counties with the Highest Percent Choosing <br> the Democratic Ballot
:::
```{r table-dem}
una_ballots %>%
  arrange(desc(pct_dem)) %>%
  head(10) %>%
  transmute(
    County       = county_desc,
    `UNA Voters` = comma(total),
    `DEM %`      = sprintf("%.1f%%", pct_dem),
    `REP %`      = sprintf("%.1f%%", pct_rep)
  ) %>%
  knitr::kable(align = "lrrr")
```

::: {style="text-align: center;"}
## Counties with the Highest Percent Choosing <br> the Republican Ballot
:::
```{r table-rep}
una_ballots %>%
  arrange(desc(pct_rep)) %>%
  head(10) %>%
  transmute(
    County       = county_desc,
    `UNA Voters` = comma(total),
    `DEM %`      = sprintf("%.1f%%", pct_dem),
    `REP %`      = sprintf("%.1f%%", pct_rep)
  ) %>%
  knitr::kable(align = "lrrr")
```

---

::: {style="text-align: center;"}
## Do Unaffiliated Voters Vote Like Their County?
:::

To test whether unaffiliated voters' ballot choices track the partisan lean of their county, I merge the 2026 primary ballot data with the 2024 presidential two-party vote share by county.

```{r trump-data}
#| cache: true

# --- Download and parse 2024 presidential results by county -----------------
# (This downloads precinct-sort files from NCSBE for all 100 counties;
#  it takes a few minutes on the first run but is cached afterward.)

library(data.table)
library(purrr)
library(janitor)

nc_county_list <- tigris::fips_codes %>%
  filter(state == "NC") %>%
  transmute(
    county_name  = str_remove(county, "\\s+County$"),
    county_upper = str_replace_all(str_to_upper(county_name), " ", "_")
  ) %>%
  distinct()

read_precinct_sort <- function(county_upper,
                               date_dir = "2024_11_05",
                               base = "https://s3.amazonaws.com/dl.ncsbe.gov/ENRS",
                               retries = 3, sleep_sec = 1) {
  url_plain  <- sprintf("%s/%s/results_precinct_sort/%s_PRECINCT_SORT.txt",
                        base, date_dir, county_upper)
  url_county <- sprintf("%s/%s/results_precinct_sort/%s_COUNTY_PRECINCT_SORT.txt",
                        base, date_dir, county_upper)
  try_download <- function(url) {
    tmp <- tempfile(fileext = ".txt")
    ok <- FALSE
    for (i in seq_len(retries)) {
      try({ utils::download.file(url, tmp, mode = "wb", quiet = TRUE)
            ok <- file.exists(tmp) && file.size(tmp) > 0 }, silent = TRUE)
      if (ok) break; Sys.sleep(sleep_sec)
    }
    if (!ok) return(NULL); tmp
  }
  f <- try_download(url_plain)
  if (is.null(f)) f <- try_download(url_county)
  if (is.null(f)) stop("Download failed for: ", county_upper)
  dt <- try(fread(f, encoding = "UTF-8", showProgress = FALSE), silent = TRUE)
  if (inherits(dt, "try-error")) dt <- fread(f, encoding = "Latin-1", showProgress = FALSE)
  janitor::clean_names(dt)
}

to_two_party_county <- function(dt_one) {
  dt_one %>%
    mutate(
      result_type_lbl     = as.character(result_type_lbl),
      candidate_party_lbl = as.character(candidate_party_lbl),
      contest_title       = as.character(contest_title),
      county              = as.character(county)
    ) %>%
    filter(result_type_lbl %in% c("STD", "Standard", "STANDARD"),
           contest_title == "US PRESIDENT") %>%
    mutate(prty = case_when(
      candidate_party_lbl %in% c("DEM", "Democratic") ~ "DEM",
      candidate_party_lbl %in% c("REP", "Republican") ~ "REP",
      TRUE ~ NA_character_
    )) %>%
    filter(!is.na(prty)) %>%
    group_by(county, prty) %>%
    summarize(votes = sum(vote_ct, na.rm = TRUE), .groups = "drop_last") %>%
    pivot_wider(names_from = prty, values_from = votes, values_fill = 0L) %>%
    ungroup() %>%
    transmute(
      county_upper    = str_replace_all(str_to_upper(county), " ", "_"),
      dem_votes       = as.numeric(DEM),
      rep_votes       = as.numeric(REP),
      total_two_party = dem_votes + rep_votes,
      trump_pct       = if_else(total_two_party > 0, 100 * rep_votes / total_two_party, NA_real_)
    )
}

county_pres <- nc_county_list$county_upper %>%
  set_names(nc_county_list$county_upper) %>%
  map(read_precinct_sort) %>%
  map(to_two_party_county) %>%
  bind_rows() %>%
  group_by(county_upper) %>%
  summarize(
    dem_votes       = sum(dem_votes, na.rm = TRUE),
    rep_votes       = sum(rep_votes, na.rm = TRUE),
    total_two_party = sum(total_two_party, na.rm = TRUE),
    trump_pct       = 100 * rep_votes / total_two_party,
    .groups = "drop"
  ) %>%
  filter(is.finite(trump_pct), total_two_party > 0)
```

```{r merge-data}
# Merge: una_ballots uses spaces in county_upper, county_pres uses underscores
scatter_df <- una_ballots %>%
  mutate(county_join = str_replace_all(county_upper, " ", "_")) %>%
  inner_join(county_pres, by = c("county_join" = "county_upper")) %>%
  mutate(harris_pct = 100 - trump_pct)

r_dem   <- cor(scatter_df$trump_pct, scatter_df$pct_dem, use = "complete.obs")
r_rep   <- cor(scatter_df$trump_pct, scatter_df$pct_rep, use = "complete.obs")
r2_dem  <- r_dem^2
r2_rep  <- r_rep^2
```

```{r scatter-theme}
theme_scatter <- function() {
  theme_minimal(base_family = "Helvetica", base_size = 12) +
    theme(
      plot.title    = element_text(face = "bold", hjust = 0.5, size = 14),
      plot.subtitle = element_text(color = "#555555", hjust = 0.5, size = 10),
      plot.caption  = element_text(color = "#888888", size = 7, hjust = 0.5),
      panel.grid.minor = element_blank()
    )
}
```

### Trump Vote Share vs. % Unaffiliated Choosing the Democratic Ballot

```{r scatter-dem}
#| fig-width: 8
#| fig-height: 6

ggplot(scatter_df, aes(x = trump_pct, y = pct_dem)) +
  geom_point(aes(size = total), alpha = 0.5, color = "#2166ac") +
  geom_smooth(method = "lm", se = TRUE, color = "#084594", linewidth = 0.8) +
  scale_size_continuous(range = c(1, 10), labels = comma, name = "UNA Voters") +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  annotate("text", x = max(scatter_df$trump_pct, na.rm = TRUE) - 2,
           y = max(scatter_df$pct_dem, na.rm = TRUE) - 2,
           label = sprintf("R² = %.3f", r2_dem),
           hjust = 1, size = 4, fontface = "italic", color = "#333333") +
  labs(
    title    = "County Trump Vote vs. Unaffiliated DEM Ballot Choice",
    subtitle = "Each point is one NC county; size = number of UNA voters who cast a ballot",
    x        = "Trump Two-Party Vote Share, 2024 (%)",
    y        = "% of Unaffiliated Voters Choosing DEM Ballot, 2026 Primary",
    caption  = "Sources: NCSBE 2024 precinct-sort results; NCSBE 2026 primary absentee data"
  ) +
  theme_scatter()
```

In counties where Trump ran strongest, unaffiliated voters overwhelmingly chose the Republican primary ballot. The relationship is strongly negative (R = -.80).

### Trump Vote Share vs. % Unaffiliated Choosing the Republican Ballot

```{r scatter-rep}
#| fig-width: 8
#| fig-height: 6

ggplot(scatter_df, aes(x = trump_pct, y = pct_rep)) +
  geom_point(aes(size = total), alpha = 0.5, color = "#b2182b") +
  geom_smooth(method = "lm", se = TRUE, color = "#a50f15", linewidth = 0.8) +
  scale_size_continuous(range = c(1, 10), labels = comma, name = "UNA Voters") +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  annotate("text", x = max(scatter_df$trump_pct, na.rm = TRUE) - 2,
           y = min(scatter_df$pct_rep, na.rm = TRUE) + 2,
           label = sprintf("R² = %.3f", r2_rep),
           hjust = 1, size = 4, fontface = "italic", color = "#333333") +
  labs(
    title    = "County Trump Vote vs. Unaffiliated REP Ballot Choice",
    subtitle = "Each point is one NC county; size = number of UNA voters who cast a ballot",
    x        = "Trump Two-Party Vote Share, 2024 (%)",
    y        = "% of Unaffiliated Voters Choosing REP Ballot, 2026 Primary",
    caption  = "Sources: NCSBE 2024 precinct-sort results; NCSBE 2026 primary absentee data"
  ) +
  theme_scatter()
```

The mirror image: unaffiliated voters in Trump-heavy counties chose the Republican ballot at very high rates (R=.80). (The two R values are identical because % DEM and % REP are complements).

### Combined View: UNA Ballot Margin vs. County Partisan Lean

```{r scatter-margin}
#| fig-width: 8
#| fig-height: 6

ggplot(scatter_df, aes(x = trump_pct - 50, y = margin)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(aes(size = total, color = margin), alpha = 0.6, show.legend = FALSE) +
  geom_smooth(method = "lm", se = TRUE, color = "#333333", linewidth = 0.8) +
  scale_color_gradient2(low = "#b2182b", mid = "#f7f7f7", high = "#2166ac", midpoint = 0) +
  scale_size_continuous(range = c(1, 10)) +
  scale_x_continuous(labels = function(x) {
    ifelse(x > 0, paste0("R+", x), ifelse(x < 0, paste0("D+", abs(x)), "Even"))
  }) +
  scale_y_continuous(labels = function(x) {
    ifelse(x > 0, paste0("D+", round(x)), ifelse(x < 0, paste0("R+", abs(round(x))), "Even"))
  }) +
  labs(
    title    = "County Partisan Lean vs. UNA Ballot Margin",
    subtitle = "Counties that lean Republican in general elections also see their UNA voters pick the GOP ballot",
    x        = "County Partisan Lean (Trump margin, 2024)",
    y        = "UNA Ballot Margin (DEM% − REP%), 2026 Primary",
    caption  = "Sources: NCSBE 2024 precinct-sort results; NCSBE 2026 primary absentee data"
  ) +
  theme_scatter()
```

This plot shows the same relationship on a margin-vs-margin scale: counties in the lower-right (Republican-leaning counties where UNA voters picked Republican ballots) and upper-left (Democratic counties where UNA voters picked Democratic ballots) confirm that unaffiliated voters broadly follow their county's partisan lean.
